---
title: "Introduction to RBPEqBind"
author: "Soon Yi"
date: "`r Sys.Date()`"
output: 
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Introduction to RBPEqBind}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

# Overview

**RBPEqBind** simulates competitive binding of multiple RNA-binding proteins (RBPs) 
to RNA sequences using an equilibrium binding model. This package provides tools for:

- Loading RBP affinity models from k-mer tables
- Simulating competitive binding at equilibrium  
- Visualizing binding profiles and competition patterns
- Exporting results to various formats including SummarizedExperiment

# Introduction

## Background

RBPEqBind implements a mathematical simulation framework for RNA-RBP interactions 
based on equilibrium binding kinetics. The simulation follows two key principles:

1. **Simultaneous competition**: All RBPs are introduced simultaneously, ensuring 
   equal chances of interaction with the RNA.
2. **Independent binding sites**: Each RNA binding site and its interactions with 
   RBPs are treated as independent events, allowing calculation of equilibrium 
   concentrations for each binding site.

## Motivation

Competitive binding among multiple RNA-binding proteins (RBPs) is a fundamental regulator of post-transcriptional gene expression. Understanding how multiple RBPs compete for overlapping binding sites is critical for interpreting transcriptome-wide binding data (such as CLIP-seq) and functional genomics studies. `RBPEqBind` provides a mathematical simulation framework to model competitive RBP binding kinetics on transcript sequences, allowing researchers to explore how RBP binding landscapes are altered by relative protein abundances, RNA concentration, and binding affinity distributions.

## Mathematical Model

For an RNA sequence of length $l$ interacting with $N$ RBPs, there exist $l - (k-1)$ 
possible k-mer binding sites. For each binding site $R$ on the RNA, the kinetic 
reaction scheme is:

$$R + P_i \rightleftharpoons P_iR$$

At equilibrium, the dissociation constant is defined as:

$$K_{D,i} = \frac{[R_{eq}][P_{i,eq}]}{[P_iR]}$$

Given the initial concentrations of RBPs ($[P_{0,i}]$) and RNA ($[R_0]$), the 
equilibrium concentration of each RNA-RBP complex can be determined by solving 
the system of equations with the constraint:

$$[R_0] \geq [R_{eq}] \geq 0$$

From the equilibrium concentrations, binding probability at each site ($p_{site}$) 
is calculated as $[P_iR] / [R_0]$. Per-position probability ($p_{pos}$) is then 
obtained by averaging overlapping $p_{site}$ values (up to $k$ overlaps per position).

For RBP enrichment analysis, fold-change is calculated by comparing the proportion 
of each RBP to a uniform distribution across the RNA sequence.

### Simulation Initial Conditions

For simulation purposes, the initial conditions are set so that all $[P_{i,0}]$ 
are in excess of the target RNA $[R_0]$. For each binding site, the effective 
initial RNA concentration needs to be determined, as a given binding site may 
exist more than once in the RNA sequence. Thus, the effective initial concentration 
of a binding site $[R_{0,e}]$ is calculated as $n \times [R_0]$, where $n$ is the 
number of times the binding site motif appears in the RNA sequence.

## Reference

For a detailed description of the mathematical model and its biological applications, 
please refer to:

> Yi S, Singh SS, Ye X, Krishna R, Kothwela V, Jankowsky E, Luna JM. (2025). 
> *Inherent Specificity and Mutational Sensitivity as Quantitative Metrics for RBP Binding.* 
> bioRxiv. https://www.biorxiv.org/content/10.1101/2025.03.28.646018v4

# Installation

To install `RBPEqBind`, start R and use the `BiocManager` package:

```r
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("RBPEqBind")
```

For the development version from GitHub, you can use `devtools` or `pak`:

```r
# install.packages("devtools")
devtools::install_github("S00NYI/RBPEqBind")
```

# Setup

```{r load-packages}
library(RBPEqBind)
library(data.table)
library(ggplot2)
```

# Data Loading

## Loading RBP Models

RBP affinity models are loaded from CSV files containing k-mer motifs and scores 
(enrichment or affinity values). The CSV file requires a column of sequence motifs 
and score columns for each RBP.

The RBP-RNA affinity scores can be derived from:

- **RNA Bind-n-Seq** (RBNS)
- **RNACompete**
- Other similar RBP target enrichment experiments

> For generating motif enrichment scores from CLIP experiments, please refer to 
> the companion R package **RBPSpecificity** (in development).

```{r load-models}
# Load raw models from package sample data
model_file <- system.file("extdata", "model_RBP.csv", package = "RBPEqBind")
raw_models <- loadModel(model_file)

# View available RBPs
names(raw_models)
```

The included four model RBPs (HH, HL, LH, LL) are designed to bind the 5-mer 
poly-U motif (UUUUU) with highest affinity, but with varying binding specificity 
and affinity distributions. For details on how these model RBP scores were 
generated, please refer to the manuscript referenced above.

## Visualizing Raw Models

The `viewModel()` function can be used to inspect the score distributions of 
raw models before affinity conversion:

```{r view-raw-models, fig.cap="Raw score distributions for model RBPs"}
viewModel(raw_models, rbp = c("HH", "HL", "LH", "LL"), bins = 100, alpha = 0.4)
```

## Setting Affinity Ranges

The included RBP models have enrichment scores normalized to a 0-to-1 scale. 
For binding simulation, each RBP may have different minimum and maximum affinity 
ranges, which are set using the `setModel()` function:

```{r set-models}
# High affinity (H) = high Ka -> low Kd
# Low affinity (L) = low Ka -> high Kd
max_affinities <- c(
  "HH" = 100, "HL" = 100,  # Ka_max = 100 -> Kd_min = 0.01 nM
  "LH" = 10,  "LL" = 10    # Ka_max = 10  -> Kd_min = 0.1 nM
)

min_affinities <- c(
  "HH" = 0.001, "HL" = 0.001,  # Ka_min = 0.001 -> Kd_max = 1000 nM
  "LH" = 0.001, "LL" = 0.001
)

rbp_models <- setModel(raw_models, 
                        max_affinity = max_affinities, 
                        min_affinity = min_affinities)
```

The `viewModel()` function can be used again to inspect the converted affinity 
(Ka) or dissociation constant (Kd) distributions:

```{r view-models-all, fig.cap="Ka distributions for all RBPs"}
viewModel(rbp_models, rbp = c("HH", "HL", "LH", "LL"), metric = "Ka", 
           bins = 100, alpha = 0.4)
```

To compare specific RBPs, pass only the desired subset:

```{r view-models-subset, fig.cap="Comparing HH vs LL affinity distributions"}
viewModel(rbp_models, rbp = c("HH", "LL"), metric = "Ka", bins = 100, alpha = 0.5)
```

# RNA-RBP Binding Simulations

Given the initial concentrations of RBPs and RNA, along with association constants 
(or dissociation constants) per motif per RBP, RBPEqBind calculates several metrics 
for each nucleotide position:

| Metric | Description |
|--------|-------------|
| `occupancy` | Binding probability (0-1) |
| `density` | Normalized binding (sum = 1) |
| `density_fc` | Fold-change over expectation |
| `occupancy_fc` | Fold-change over mean occupancy |

**For single RBP simulations** (only one RBP with concentration > 0):

- `density_fc = density × L` (enrichment over uniform distribution)
- `occupancy_fc = occupancy / mean(occupancy)`

**For multiple active RBPs**:

- `density_fc = self_density / sum(other_densities)`
- `occupancy_fc = self_occupancy / sum(other_occupancies)`

This vignette covers four simulation scenarios of increasing complexity:

1. **Scenario 1**: Single sequence, fixed concentrations
2. **Scenario 2**: Single sequence, concentration grid sweep
3. **Scenario 3**: FASTA file, fixed concentrations
4. **Scenario 4**: FASTA file, concentration grid sweep

# Scenario 1: RBPs Binding to a Single RNA Target at Fixed Concentrations

## Generating or Loading RNA Sequence

Users can provide their own sequence or use the built-in `generateRNA()` function 
to create randomized RNA sequences:

```{r generate-rna}
# Generate a random 100-nt RNA sequence
random_seq <- generateRNA(100)
cat("Generated sequence:", substr(random_seq, 1, 50), "...\n")

# Or use a specific reference sequence
seq <- "UAGCGGUGCGAUUGGCCCGUGGACCGCGUUUUUGCACUCAUCGUUUCGCACUAAGUACAUAUAGUUGCGACAAAGCCGCUUAUGAGUUGGGGGUAUAUUC"
```

## Running the Simulation

Set initial concentrations of RBPs and target RNA, then run the simulation:

```{r scenario1-simulate}
# Define concentrations (nM) - competitive binding of all 4 RBPs
prot_concs <- c("HH" = 100, "HL" = 100, "LH" = 100, "LL" = 100)
rna_conc <- 10.0  # nM

# Detect k-mer size from models
k_size <- nchar(rbp_models$HH$motif[1])

# Run simulation
res <- simulateBinding(
  sequence = seq,
  rbp_models = rbp_models,
  protein_concs = prot_concs,
  rna_conc = rna_conc,
  k = k_size
)

# Add transcript column for visualization
res$transcript <- "Custom RNA"

head(res)
```

## Visualization

Each metric calculated using `simulateBinding()` can be visualized with built-in 
plotting functions.

### Per-Position Binding Profile

The `plotBinding()` function displays binding metrics across positions:

> **Note:** The visualizations below show a zoomed-in view from positions 20-60. 
> The `xlim` parameter can be removed to show the entire sequence, or adjusted 
> to view a specific segment of the target RNA.

```{r scenario1-binding-raw, fig.cap="Raw density_fc (positions 20-60)"}
plotBinding(res, rbp = c("HH", "HL", "LH", "LL"), metric = "density_fc", 
             transcript = "Custom RNA", xlim = c(20, 60))
```

For smoother visualization, apply a k-mer sliding window average:

```{r scenario1-binding-window, fig.cap="Density FC with 5-mer smoothing window (positions 20-60)"}
plotBinding(res, rbp = c("HH", "HL", "LH", "LL"), metric = "density_fc", 
             transcript = "Custom RNA", window = 5, xlim = c(20, 60))
```

### Heatmap Visualization

The `plotHeatmap()` function displays binding for multiple RBPs simultaneously:

```{r scenario1-heatmap, fig.cap="Heatmap of occupancy_fc (zoomed region)"}
plotHeatmap(res, transcript = "Custom RNA", xlim = c(20, 60), 
             xaxis_type = "both", metric = "occupancy_fc")
```

## Exporting Results

Simulation results can be exported using the built-in `exportResults()` function 
in JSON or CSV format:

```{r scenario1-export}
# Export to JSON
tmp_json <- tempfile(fileext = ".json")
exportResults(res, output_file = tmp_json, format = "json")

# Export to CSV
tmp_csv <- tempfile(fileext = ".csv")
exportResults(res, output_file = tmp_csv, format = "csv")

# Clean up
unlink(tmp_json)
unlink(tmp_csv)
```

Results can also be converted to SummarizedExperiment format:

```{r scenario1-se}
se <- makeSE(res, rbp_models = rbp_models)
se
```

# Scenario 2: RBP Binding with Varying Concentration Grid

To run simulations across multiple combinations of RBP and RNA concentrations, 
use the `simulateGrid()` function. This function includes a `parallel` option 
for increased computational efficiency.

## Running Grid Simulation

```{r scenario2-grid}
# Define concentration grids
prot_grid <- list(
  HH = c(10, 100), 
  HL = c(10, 100),
  LH = c(10, 100),
  LL = c(10, 100)
)
rna_grid <- c(10)

res_grid <- simulateGrid(
  sequence = seq,
  rbp_models = rbp_models,
  protein_conc_grid = prot_grid,
  rna_conc_grid = rna_grid,
  k = k_size,
  parallel = FALSE
)
```

## Visualization

Grid sweep results can be visualized for specific concentration combinations 
using the same plotting functions:

```{r scenario2-binding}
# Filter to specific concentration combination
plotBinding(
  results = res_grid,
  rbp = c("HH", "HL", "LH", "LL"),
  rna_conc = 10,
  protein_conc = c(HH=10, HL=10, LH=100, LL=100),
  metric = "density"
)
```

### Competition Grid Plot

The `plotGrid()` function focuses on changes in binding for one RBP across 
different target RNA and competitor RBP concentrations:

```{r scenario2-plot-grid}
# First run a 2-RBP simulation for cleaner visualization
rbp_models_2rbp <- rbp_models[c("HH", "LL")]

res_grid_2rbp <- simulateGrid(
  sequence = seq,
  rbp_models = rbp_models_2rbp,
  protein_conc_grid = list(HH = c(10, 50), LL = c(10, 50)),
  rna_conc_grid = c(10),
  k = k_size,
  parallel = FALSE
)

plotGrid(
  results = res_grid_2rbp,
  rbp1 = "HH",
  rbp2 = "LL",
  rbp1_concs = c(10, 50),
  roi_range = c(29, 33),
  metric = "density"
)
```

Grid sweep results can also be exported in JSON/CSV or SummarizedExperiment format.

# Scenario 3: RBPs Binding to Multiple Target RNAs

To simulate RBP binding to multiple RNA sequences (e.g., transcript sequences 
from bioMart), RBPEqBind provides functions to load FASTA files directly. Note that 
binding simulation is performed on a sequence-by-sequence basis, not simultaneously 
on all RNA sequences in the FASTA file.

```{r scenario3-simulate}
# Load sample FASTA from package
fasta_file <- system.file("extdata", "test_transcripts.fa", package = "RBPEqBind")

res_fasta <- simulateBindingFasta(
  fasta_file = fasta_file,
  rbp_models = rbp_models,
  protein_concs = c(HH = 100, HL = 100, LH = 100, LL = 100),
  rna_conc = 10,
  k = 5,
  parallel = FALSE
)
head(res_fasta)
```

Visualization and result export are performed the same way as in Scenario 1.

# Scenario 4: RBPs Binding to Multiple Target RNAs Across Different Concentrations

RBPEqBind also supports concentration grid sweeps across sequences in a FASTA file:

```{r scenario4-grid}
fasta_file <- system.file("extdata", "test_transcripts.fa", package = "RBPEqBind")

res_grid_fasta <- simulateGridFasta(
  fasta_file = fasta_file,
  rbp_models = rbp_models,
  protein_conc_grid = list(
    HH = c(10, 100),
    HL = c(10, 100),
    LH = c(10, 100),
    LL = c(10, 100)
  ),
  rna_conc_grid = c(10),
  k = 5,
  parallel = FALSE
)
head(res_grid_fasta)
```

Visualization and export functions work the same as in Scenario 2.

# Session Info

```{r session}
sessionInfo()
```
